---
title: "R Notebook"
output: html_notebook
---

```{r}
# load some packages
library(tidyverse)
library(ismev)
library(fExtremes)
```


# Introduction to extreme value analysis in R

We are going to jump right in with an example to illustrate a classical approach to understanding **block maxima**: maximum values within some block (e.g., a year), where we have multiple blocks.

## Annual maximum sea levels at Port Pirie, South Australia

We are going to work with yearly maximum sea levels recorded at Port Pirie, South Australia over a 64 year record from 1923 to 1987. 
Let's start off by having a look at the data. 

```{r}
data(portpirie)
str(portpirie)
```

So, we are working with a `data.frame` object, with two columns: `Year` (a number) and `SeaLevel` (also a number). 
What do the annual maxima look like over time? 

```{r}
ggplot(portpirie, aes(Year, SeaLevel)) +
  geom_line() +
  geom_point()
```

Let's say that we wanted to know the probability of having an annual maximum over 4.5. 
What's a simple way to estimate this from the data?

```{r}
# Your code here - find the probability of an annual maximum > 4.5
```


Now, let's ask a harder question: what's the probability of having an annual maximum over 5?
If we only considered the historical record, a naive approach would be to conclude that since there are no observations over 5, that the probability is 0. 
Clearly, this answer would not be sufficient if we were gambling the fate of an oceanside community on our estimate. 

Let's get a little more formal with setting up this problem. 
If we define $X$ as an annual maximum, what we want to know is the probability that $X$ is greater than 5:

$$\text{Pr}(X > 5) = ?$$

You might remember from statistics that if we know the distribution of a random variable, we can calculate the probability that the random variable takes on a value greater than some value by computing the area under the distribution curve to the right of some value (cue traumatic flashbacks to introductory statistics course), a common example being the tail area probability of a standard normal distribution: 

```{r}
q <- qnorm(.975)
xseq <- seq(q, 4, 0.01)
x <- c(q, xseq, 4)
y <- c(0, dnorm(xseq), 0) 
curve(dnorm(x), xlim=c(-4, 4), main='Standard normal distribution') 
polygon(x, y, col = 'skyblue')
text(3, 
     .05, 
     labels = paste0('Area right of ', 
                     round(q, 3), 
                     ':\n', 
                     round(1 - pnorm(q), 3)))
abline(v = q, lty = 2)
```

For the standard normal distribution, $\text{Pr}(x \ge 1.96) \approx 0.025$
If we knew the distribution of sea level maxima, we could calculate this tail area probability to get the answer to our original question ($\text{Pr}(X > 5)$). 

Unfortunately, we don't know the distribution of the annual sea level maxima. 
However, we can use the **generalized extreme value (GEV) distribution** as an approximation. 
The GEV is a good guess for the distribution based on some theoretical results, and has three parameters: location $\mu$, scale $\sigma$ (which must be $>0$), and shape $\xi$. 

The GEV distribution is interesting in part because it can do some funky things depending on the parameter values. 
For instance, it can be totally unbounded, generating values along the entire real line: 

```{r}
curve(dgev, xlim = c(-10, 10), ylab = 'GEV density')
title(expression(list(xi == 1, mu == 0, sigma == 1)))
```

